An Improved Method for Coupling Hydrodynamics with Astrophysical Reaction Networks

### 2021-12-28 simplified_sdc_intro/paper.tex
more updates

\section{Simulations}

@@

\cite{castro_sdc} introduced a test problem for measuring convergence
of a reacting hydrodynamic algorithm.  We run that same test here with the
simplified SDC algorithm.

Also explore: PLM vs PPM, 1 vs. 2 vs. 3 iterations

% run on 04-06-21
% Castro       git describe: 21.04-dirty
% AMReX        git describe: 21.03-58-g87c81d0ca
% Microphysics git describe: 21.04-1-g33a4d0fe

\begin{deluxetable}{lllllll}
\tablecaption{\label{table:react_convert_strang} Convergence ($L_1$ norm) for the reacting convergence problem using Strang splitting.}
\tablehead{\colhead{field} & \colhead{$\epsilon_{64 \rightarrow 128}$} & 
           \colhead{rate} & \colhead{$\epsilon_{128\rightarrow 256}$} & 
           \colhead{rate} & \colhead{$\epsilon_{256\rightarrow 512}$}}
\startdata
 $\rho$                      & $2.794 \times 10^{18}$  & 2.044  & $6.777 \times 10^{17}$  & 2.554  & $1.154 \times 10^{17}$  \\
 $\rho u$                    & $6.796 \times 10^{26}$  & 2.448  & $1.245 \times 10^{26}$  & 2.889  & $1.681 \times 10^{25}$  \\
 $\rho v$                    & $6.796 \times 10^{26}$  & 2.448  & $1.245 \times 10^{26}$  & 2.889  & $1.681 \times 10^{25}$  \\
 $\rho E$                    & $2.451 \times 10^{35}$  & 2.351  & $4.803 \times 10^{34}$  & 2.742  & $7.179 \times 10^{33}$  \\
 $\rho e$                    & $2.261 \times 10^{35}$  & 2.320  & $4.526 \times 10^{34}$  & 2.821  & $6.403 \times 10^{33}$  \\
 $T$                         & $2.237 \times 10^{21}$  & 1.691  & $6.927 \times 10^{20}$  & 2.482  & $1.240 \times 10^{20}$  \\
 $\rho X(\isotm{He}{4})$     & $2.878 \times 10^{18}$  & 2.020  & $7.096 \times 10^{17}$  & 2.529  & $1.229 \times 10^{17}$  \\
 $\rho X(\isotm{C}{12})$     & $1.698 \times 10^{17}$  & 1.950  & $4.393 \times 10^{16}$  & 2.232  & $9.353 \times 10^{15}$  \\
 $\rho X(\isotm{O}{16})$     & $1.687 \times 10^{14}$  & 1.660  & $5.338 \times 10^{13}$  & 1.957  & $1.375 \times 10^{13}$  \\
 $\rho X(\isotm{Fe}{56})$    & $2.794 \times 10^{8}$   & 2.044  & $6.777 \times 10^{7}$   & 2.554  & $1.154 \times 10^{7}$   \\
\enddata
\end{deluxetable}

% run on 04-06-21
% Castro       git describe: 21.04-dirty
% AMReX        git describe: 21.03-58-g87c81d0ca
% Microphysics git describe: 21.04-1-g33a4d0fe

\begin{deluxetable}{lllllll}
\tablecaption{\label{table:react_convert_sdc} Convergence ($L_1$ norm) for the reacting convergence problem using SDC integration (2 iterations).}
\tablehead{\colhead{field} & \colhead{$\epsilon_{64 \rightarrow 128}$} & 
           \colhead{rate} & \colhead{$\epsilon_{128\rightarrow 256}$} & 
           \colhead{rate} & \colhead{$\epsilon_{256\rightarrow 512}$}}
\startdata
 $\rho$                      & $2.795 \times 10^{18}$  & 2.044  & $6.777 \times 10^{17}$  & 2.554  & $1.154 \times 10^{17}$  \\
 $\rho u$                    & $6.798 \times 10^{26}$  & 2.449  & $1.245 \times 10^{26}$  & 2.889  & $1.680 \times 10^{25}$  \\
 $\rho v$                    & $6.798 \times 10^{26}$  & 2.449  & $1.245 \times 10^{26}$  & 2.889  & $1.680 \times 10^{25}$  \\
 $\rho E$                    & $2.452 \times 10^{35}$  & 2.352  & $4.804 \times 10^{34}$  & 2.742  & $7.179 \times 10^{33}$  \\
 $\rho e$                    & $2.262 \times 10^{35}$  & 2.321  & $4.527 \times 10^{34}$  & 2.822  & $6.404 \times 10^{33}$  \\
 $T$                         & $2.236 \times 10^{21}$  & 1.691  & $6.927 \times 10^{20}$  & 2.481  & $1.240 \times 10^{20}$  \\
 $\rho X(\isotm{He}{4})$     & $2.879 \times 10^{18}$  & 2.020  & $7.096 \times 10^{17}$  & 2.529  & $1.229 \times 10^{17}$  \\
 $\rho X(\isotm{C}{12})$     & $1.697 \times 10^{17}$  & 1.950  & $4.393 \times 10^{16}$  & 2.231  & $9.357 \times 10^{15}$  \\
 $\rho X(\isotm{O}{16})$     & $1.686 \times 10^{14}$  & 1.659  & $5.338 \times 10^{13}$  & 1.957  & $1.375 \times 10^{13}$  \\
 $\rho X(\isotm{Fe}{56})$    & $2.795 \times 10^{8}$   & 2.044  & $6.777 \times 10^{7}$   & 2.554  & $1.154 \times 10^{7}$   \\
\enddata
\end{deluxetable}

\subsection{X-ray Burst}

We next consider a flame propagating through an accreted helium layer on a neutron star,
as described in \cite{harpole:2021}.

@@

We use a network with 28 nuclei and 96 rates that captures He burning
with links for $(\alpha,p)(p,\gamma)$ reactions and the rates
involving \isot{N}{14} to bypass
$\isotm{C}{12}(\alpha,\gamma)\isotm{O}{16}$ as described in
\cite{shenbildsten}.  The rates are taken from the ReacLib library
\citep{reaclib} and the network is written by
\pynucastro~\citep{pynucastro}.  The nuclei and links in the network are
shown in Figure~\ref{fig:subch_network}.  The many reaction pathways in this network
make it a good test for the SDC algorithm.

The initial model is generated following the procedure described in \citep{subchandra}.
For the present work, we are only concerned with the behavior of the time-integration strategy
and not on the astrophysical implications of the physical setup we explore.

@@

We do a suite of runs with both Strang and the simplified-SDC
integration, using CFL numbers of 0.2, 0.4, and 0.8 and reaction
network tolerances of $\atolspec = \rtolspec = 10^{-5}$ and $\atolener
= \rtolener = 10^{-6}$.  We also run two sets with tighter tolerances.
The runs labeled ``mixed\_tols'' have $\atolspec = 10^{-8}$ and the runs
labeled ``tol1.e-6'' have $\atolspec = \rtolspec = 10^{-6}$, with the other
tolerances unchanged.

@@

>>>>

\section{Tests}

@@

\cite{castro_sdc} introduced a test problem for measuring convergence
of a reacting hydrodynamic algorithm.  This is based on the acoustic
pulse problem from \citep{mccorquodalecolella}, and was also used in
\citep{strang_rnaas}.  A periodic domain with a uniform entropy is
initialized with a pressure profile with a small perturbation.  This
drives a low amplitude acoustic wave radially outward from the center.
A simple reaction network with the triple-alpha and
$\isotm{C}{12}(\alpha,\gamma)\isotm{O}{16}$ rates releases energy.  By
running at different resolutions, with a fixed timestep that scales
with resolution, we can compute the convergence rate.  In order to
demonstrate second-order convergence, the problem must be smooth (so
the slope limiters don't kick in severely), which means that the
energy release cannot be so vigorous as to drive shocks.  test here
with the simplified SDC algorithm.

% EOS: /home/zingale/development/Microphysics/EOS/helmholtz
% NETWORK: /home/zingale/development/Microphysics/networks/triple_alpha_plus_cago

% Castro       git describe: 21.12-9-gd7a27ccd5
% AMReX        git describe: 21.12-64-gfde72340e
% Microphysics git describe: 21.12-12-g9595a34f

\begin{deluxetable}{lllllll}
\tablecaption{\label{table:react_converge_strang} Convergence ($L_1$ norm) for the reacting convergence problem using Strang splitting.}
\tablehead{\colhead{field} & \colhead{$\epsilon_{64 \rightarrow 128}$} & 
           \colhead{rate} & \colhead{$\epsilon_{128\rightarrow 256}$} & 
           \colhead{rate} & \colhead{$\epsilon_{256\rightarrow 512}$}}
\startdata
 $\rho$                      & $2.794 \times 10^{18}$  & 2.044  & $6.777 \times 10^{17}$  & 2.554  & $1.154 \times 10^{17}$  \\
 $\rho u$                    & $6.796 \times 10^{26}$  & 2.448  & $1.245 \times 10^{26}$  & 2.889  & $1.681 \times 10^{25}$  \\
 $\rho v$                    & $6.796 \times 10^{26}$  & 2.448  & $1.245 \times 10^{26}$  & 2.889  & $1.681 \times 10^{25}$  \\
 $\rho E$                    & $2.451 \times 10^{35}$  & 2.351  & $4.803 \times 10^{34}$  & 2.742  & $7.179 \times 10^{33}$  \\
 $\rho e$                    & $2.261 \times 10^{35}$  & 2.320  & $4.526 \times 10^{34}$  & 2.821  & $6.403 \times 10^{33}$  \\
 $T$                         & $2.237 \times 10^{21}$  & 1.691  & $6.927 \times 10^{20}$  & 2.482  & $1.240 \times 10^{20}$  \\
 $\rho X(\isotm{He}{4})$     & $2.878 \times 10^{18}$  & 2.020  & $7.096 \times 10^{17}$  & 2.529  & $1.229 \times 10^{17}$  \\
 $\rho X(\isotm{C}{12})$     & $1.698 \times 10^{17}$  & 1.950  & $4.393 \times 10^{16}$  & 2.232  & $9.353 \times 10^{15}$  \\
 $\rho X(\isotm{O}{16})$     & $1.687 \times 10^{14}$  & 1.660  & $5.338 \times 10^{13}$  & 1.957  & $1.375 \times 10^{13}$  \\
 $\rho X(\isotm{Fe}{56})$    & $2.794 \times 10^{8}$   & 2.044  & $6.777 \times 10^{7}$   & 2.554  & $1.154 \times 10^{7}$   \\
\enddata
\end{deluxetable}

\begin{deluxetable}{lllllll}
\tablecaption{\label{table:react_converge_sdc} Convergence ($L_1$ norm) for the reacting convergence problem using SDC integration (2 iterations).}
\tablehead{\colhead{field} & \colhead{$\epsilon_{64 \rightarrow 128}$} & 
           \colhead{rate} & \colhead{$\epsilon_{128\rightarrow 256}$} & 
           \colhead{rate} & \colhead{$\epsilon_{256\rightarrow 512}$}}
\startdata
 $\rho$                      & $2.784 \times 10^{18}$  & 2.048  & $6.734 \times 10^{17}$  & 2.558  & $1.143 \times 10^{17}$  \\
 $\rho u$                    & $6.779 \times 10^{26}$  & 2.447  & $1.243 \times 10^{26}$  & 2.895  & $1.671 \times 10^{25}$  \\
 $\rho v$                    & $6.779 \times 10^{26}$  & 2.447  & $1.243 \times 10^{26}$  & 2.895  & $1.671 \times 10^{25}$  \\
 $\rho E$                    & $2.450 \times 10^{35}$  & 2.353  & $4.795 \times 10^{34}$  & 2.737  & $7.193 \times 10^{33}$  \\
 $\rho e$                    & $2.256 \times 10^{35}$  & 2.320  & $4.518 \times 10^{34}$  & 2.817  & $6.414 \times 10^{33}$  \\
 $T$                         & $2.232 \times 10^{21}$  & 1.695  & $6.893 \times 10^{20}$  & 2.484  & $1.233 \times 10^{20}$  \\
 $\rho X(\isotm{He}{4})$     & $2.866 \times 10^{18}$  & 2.024  & $7.049 \times 10^{17}$  & 2.534  & $1.217 \times 10^{17}$  \\
 $\rho X(\isotm{C}{12})$     & $1.695 \times 10^{17}$  & 1.959  & $4.357 \times 10^{16}$  & 2.236  & $9.250 \times 10^{15}$  \\
 $\rho X(\isotm{O}{16})$     & $1.681 \times 10^{14}$  & 1.662  & $5.313 \times 10^{13}$  & 1.963  & $1.363 \times 10^{13}$  \\
 $\rho X(\isotm{Fe}{56})$    & $2.784 \times 10^{8}$   & 2.048  & $6.734 \times 10^{7}$   & 2.558  & $1.143 \times 10^{7}$   \\
\enddata
\end{deluxetable}

Tables \ref{table:react_converge_strang} and
\ref{table:react_converge_sdc} show the Strang and simplified-SDC
convergence respectively.  We see that both methods converge nearly
second order, and agree well with each other.

@@

We next model a double detonation Type Ia
supernova.  In this model, a detonation begins in the accreted helium
layer on a sub-Chandrasekhar mass white dwarf and that detonation can
either drive a compression wave into the core igniting a second
detonation in the carbon / oxygen core, or burn through the interface
at the base of the accreted layer producing an inward propagating
carbon detonation.  Our goal here is not to explore the feasibility of
the model or understand whether a detonation propagating through the
He-C interface is physical, but rather to look at the coupling of the
hydrodynamics and reactions.  For that reason, we explore only a
single model and start off with a rather large temperature
perturbation in the He layer.

We use a network with 28 nuclei and 96 rates that captures He burning
with links for $(\alpha,p)(p,\gamma)$ reactions and the rates
involving \isot{N}{14} to bypass
$\isotm{C}{12}(\alpha,\gamma)\isotm{O}{16}$ as described in
\cite{shenbildsten}.  The rates are taken from the ReacLib library
\citep{reaclib} and the network is written by
\pynucastro~\citep{pynucastro} directly in the C++ format that our
code requires.  The nuclei and links in the network are shown in
Figure~\ref{fig:subch_network}.  This moderate-sized network with many
reaction pathways (forward and reverse) make it a good test for the
coupling between hydro and reactions.  An overview of the network is
shown in Figure~\ref{fig:subch_network}.

@@

The initial model is generated following the procedure described in
\citep{subchandra}.  We use a $1.1~M_\odot$ carbon white dwarf with a
$0.05~M_\odot$ He layer.  We seed the He layer with 1\% (by mass) of
\isot{N}{14}.  The white dwarf is isothermal and the temperature ramps
up at the base of the He layer and is then isentropic until the
surface.  Figure~\ref{fig:subch_initial_model} shows the structure of
the initial model.  We use 2d axisymmetric coordinates with a domain
size of $1.024\times 10^9~\mathrm{cm} \times 2.048\times
10^9~\mathrm{cm}$, a base grid of $256 \times 512$ zones and two
levels of refinement (both jumping by $2\times$), giving a fine-grid
resolution of $10~\mathrm{km}$.  Gravity is modeled as a monopole.

\begin{figure}[t]
\centering
\plotone{initial_model}
\caption{\label{fig:subch_initial_model} Initial model for the double detonation test problem.}
\end{figure}

An initial perturbation is placed on the symmetry axis of the form:

We do a suite of runs with both Strang and the simplified-SDC
integration, using CFL numbers of 0.2, 0.4, and 0.8 and reaction
network tolerances of $\atolspec = \rtolspec = 10^{-5}$ and $\atolener
= \rtolener = 10^{-6}$.  We also run two sets with tighter tolerances.
The runs labeled ``mixed\_tols'' have $\atolspec = 10^{-8}$ and the
runs labeled ``tol1.e-6'' have $\atolspec = \rtolspec = 10^{-6}$, with
the other tolerances unchanged.  For the step rejection part of VODE,
we use $f = 4$ and $\eta = 4$.

@@

We have not focused on performance or load-balancing the reactions in
the tests that we ran, but these limited runs show that the
simplified-SDC method runs as fast (or faster) than the Strang method
for the same tolerances and CFL, despite having to do the advection
twice.  We would expect similar behavior for any simulations that are
dominated by the cost of reactions.
